---
title: Test Post
author: Robert A. Amezquita
date: '2017-04-04'
slug: ''
categories: []
tags: []
description: ''
featured: ''
featuredalt: ''
featuredpath: ''
linktitle: ''
type: 'post'
---

Trying out a new post just to see how this system works. Already found that (featured) images don't seem to be working on the home page, but do work on the /blog/ subdomain. Also I need to make sure to set the "type: 'post'" in the YAML header. Anyways, let's see how beautifully R code renders..

```{r setup, include = FALSE}

library(tidyverse)

```

```{r test, echo = TRUE}

## run in a chunk with `include = FALSE`
## library(tidyverse) 

## Run within this chunk with echo = TRUE
hello_world <- frame_data(
    ~first, ~second, ~third, ~fourth,
    "its", "a", "beautiful", "day"
)

hello_world

```

Now what happens when I generate a plot?


```{r test2, fig.width = 2, fig.height = 2}

frame_data(
    ~one, ~two,
    1, 2,
    1, 2
) %>%
    ggplot(aes(x = one, y = two)) +
    geom_point() +
    theme_classic()
    
```

Looking beautiful!! And yes, I am a `theme_classic()` kind of person myself, though I also enjoy using the `cowplot` package for really nice plot setting. 
